---
title: "Independent Reanalysis: COVID/Shingles Vaccine Signals Amid FDA Publication Blocks"
subtitle: "Bayesian Disproportionality Across mRNA, Viral-Vector, and Zoster Products in VAERS"
author: "Global Patient Safety"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    code-fold: true
    code-summary: "Show code"
    theme: flatly
    fig-width: 10
    fig-height: 6
    self-contained: true
execute:
  echo: true
  warning: false
  message: false
  freeze: auto
---

```{r setup}
library(arrow)
library(dplyr)
library(tidyr)
library(ggplot2)
library(forcats)
library(gt)
library(scales)
library(stringr)

# Resolve the VAERS signals parquet from a list of candidate paths so this
# QMD renders both on developer workstations and on the VPS.
# Need the FULL parquet (with prr/ror/ic/observed/etc.), not the 4-column
# aggregated splash version that powers the gps-patient app's headline.
SIGNALS_CANDIDATES <- c(
  "/srv/shiny-server/gps-patient/data/signals_vaers_full.parquet",
  "/home/harlan/data/signal-compute/vaers/signals_vaers_v2026-05-03.parquet"
)
SIGNALS_PATH <- SIGNALS_CANDIDATES[file.exists(SIGNALS_CANDIDATES) |
                                   dir.exists(SIGNALS_CANDIDATES)][1]
stopifnot(!is.na(SIGNALS_PATH))

vaers_raw_quarterly <- open_dataset(SIGNALS_PATH) |> collect()

# The full parquet stores ONE ROW PER (drug, event, quarter) — the
# disproportionality stats recomputed at each quarterly snapshot.
#
# For each (drug, event) pair we keep the FIRST quarter at which it
# crossed the project's consensus signal threshold (EB05 > 2 AND
# flagged by >= 2 of 4 methods). This is the regulatory-relevant moment:
# "when would real-time surveillance have first detected this?". Stats
# attached to the row are exactly those at first detection — not the
# peak, not the latest, not an average across quarters. This is the
# same convention the existing COVID vaccine article uses via the
# pre-computed covid_first_seen.parquet table.
vaers_raw <- vaers_raw_quarterly |>
  filter(eb05 > 2, n_methods_flagged >= 2) |>     # signal-grade rows only
  group_by(drug, event) |>
  arrange(quarter, .by_group = TRUE) |>
  slice_head(n = 1) |>                            # earliest signaling quarter
  ungroup() |>
  rename(first_quarter = quarter)

cat("After collapse to first-detection quarter:",
    format(nrow(vaers_raw), big.mark = ","),
    "unique (drug, event) signals\n")
```

## Context

This analysis is prompted by recent reports of delays and blocks on the
publication of vaccine safety studies, including FDA-funded analyses, in
May 2026. Because the underlying spontaneous-reporting data — CDC's VAERS —
remains publicly accessible, an independent reanalysis using standard
pharmacovigilance methods is feasible regardless of the publication
status of any particular paper.

This article applies the same disproportionality framework used elsewhere
on the site — GPS / PRR / ROR / BCPNN-IC, with a multi-method consensus
threshold of "flagged by at least 2 of 4 methods" — to two vaccine
families: **COVID-19 vaccines** (nine products) and **shingles
vaccines** (three products: Shingrix recombinant, Zostavax live, and
unbranded Zoster reports). The focus is on three pre-specified event
domains: **cardiac**, **neurological**, and **thrombotic** outcomes.
Pre-specifying systems of interest narrows multiple-comparisons exposure
and matches the prompt for this reanalysis.

::: callout-note
**This is a signal-detection reanalysis, not a causal study.**
Disproportionality methods identify pairs where the reporting rate is
unusually high relative to the rest of VAERS. Such signals are
hypotheses, not evidence of causation. Confirmation requires controlled
study designs, denominator-aware cohort or self-controlled methods, and
clinical adjudication.
:::

## Methods

All four methods come from the `safetysignal` R package developed for
this project (see project-wide methodology notes). For each
`(vaccine, event)` pair counted in VAERS:

- **GPS (Gamma-Poisson Shrinker)** — DuMouchel 1999 dual-Gamma posterior.
  `EB05` is the 5th-percentile credible lower bound on the linear
  relative-reporting-ratio scale (a direct posterior quantile, **not** the
  EBGM geometric mean which would apply a log transformation).
- **PRR (Proportional Reporting Ratio)** — frequentist relative
  reporting rate with χ² test. Threshold: PRR > 2, χ² > 4, n ≥ 3.
- **ROR (Reporting Odds Ratio)** — odds-ratio formulation of the same
  2×2. Threshold: lower 95% CI > 1.
- **BCPNN/IC (Information Component)** — Bayesian shrinkage; IC₀₂₅ > 0
  to flag.

A pair is treated as a candidate signal if **EB05 > 2 AND it is flagged
by ≥ 2 of the 4 methods**.

```{r classifiers}
label_covid_vaccine <- function(drug) {
  case_when(
    grepl("PFIZER.*BIVALENT|BIVALENT.*PFIZER", drug, ignore.case = TRUE) ~ "Pfizer Bivalent",
    grepl("MODERNA.*BIVALENT|BIVALENT.*MODERNA", drug, ignore.case = TRUE) ~ "Moderna Bivalent",
    grepl("PFIZER|BIONTECH|COMIRNATY",           drug, ignore.case = TRUE) ~ "Pfizer-BioNTech",
    grepl("MODERNA|SPIKEVAX|MNEXSPIKE",           drug, ignore.case = TRUE) ~ "Moderna",
    grepl("JANSSEN|JOHNSON",                      drug, ignore.case = TRUE) ~ "Janssen (J&J)",
    grepl("NOVAVAX",                              drug, ignore.case = TRUE) ~ "Novavax",
    TRUE ~ "Other / unknown"
  )
}

label_shingles_vaccine <- function(drug) {
  case_when(
    grepl("SHINGRIX",  drug, ignore.case = TRUE) ~ "Shingrix (recombinant)",
    grepl("ZOSTAVAX",  drug, ignore.case = TRUE) ~ "Zostavax (live)",
    grepl("ZOSTER",    drug, ignore.case = TRUE) ~ "Zoster (no brand)",
    TRUE ~ "Other"
  )
}

# Three pre-specified event domains. Patterns are intentionally broad so
# that closely related PTs (e.g., 'myocardial infarction' vs 'acute
# myocardial infarction') collapse into the same domain row.
classify_event <- function(event) {
  case_when(
    grepl("myocard|pericard|cardiac|heart|arrhyth|fibrillat|tachycard|infarct|conduction",
          event, ignore.case = TRUE) ~ "Cardiac",
    grepl("thrombo|embol|clot|coagul|disseminated|purpura|platelet|heparin|anti-platelet",
          event, ignore.case = TRUE) ~ "Thrombotic",
    grepl("neuro|encephal|guillain|myelitis|neurop|seizure|convuls|paraly|bell|paresthes|tremor|ataxia|aphasia",
          event, ignore.case = TRUE) ~ "Neurological",
    TRUE ~ NA_character_
  )
}

covid <- vaers_raw |>
  filter(grepl("COVID|SARS|BNT|MRNA|PFIZER|MODERNA|JANSSEN|NOVAVAX", drug, ignore.case = TRUE)) |>
  mutate(vaccine = label_covid_vaccine(drug), family = "COVID-19")

shingles <- vaers_raw |>
  filter(grepl("ZOSTER|SHINGRIX|ZOSTAVAX", drug, ignore.case = TRUE)) |>
  mutate(vaccine = label_shingles_vaccine(drug), family = "Shingles")

both <- bind_rows(covid, shingles) |>
  mutate(category = classify_event(event)) |>
  filter(!is.na(category))
```

## Headline counts

Each row of every table that follows is **a unique (vaccine, event) signal**
counted once at its first-detection quarter. "Signals detected" therefore
counts distinct pairs that crossed the consensus threshold at some point in
the data window, not the number of quarterly observations.

```{r headline}
both |>
  group_by(family, vaccine) |>
  summarise(
    signals_detected  = n(),
    signals_4_methods = sum(n_methods_flagged == 4, na.rm = TRUE),
    max_eb05          = round(max(eb05, na.rm = TRUE), 1),
    earliest_signal   = min(first_quarter, na.rm = TRUE),
    .groups = "drop"
  ) |>
  arrange(family, desc(signals_detected)) |>
  gt() |>
  tab_header(title = "Distinct signals per vaccine — cardiac, neurological, thrombotic only",
             subtitle = "One row per (vaccine, event) at its first-detection quarter") |>
  fmt_number(columns = c(signals_detected, signals_4_methods), decimals = 0) |>
  cols_label(
    family            = "Family",
    vaccine           = "Vaccine",
    signals_detected  = "Signals detected",
    signals_4_methods = "Flagged by all 4 methods",
    max_eb05          = "Max EB05",
    earliest_signal   = "Earliest"
  ) |>
  tab_options(table.font.size = 13)
```

## Top signals — cardiac

```{r cardiac}
cardiac_top <- both |>
  filter(category == "Cardiac") |>
  arrange(desc(eb05)) |>
  group_by(family) |>
  slice_head(n = 12) |>
  ungroup()

cardiac_top |>
  select(family, vaccine, event, first_quarter, eb05, prr, ic, n_methods_flagged) |>
  gt() |>
  tab_header(title = "Top cardiac signals — COVID-19 + shingles vaccines",
             subtitle = "Statistics at first-detection quarter · top 12 per family by EB05") |>
  fmt_number(columns = c(eb05, prr, ic), decimals = 1) |>
  data_color(columns = eb05,
             palette = c("white", "#a50f15"),
             alpha = 0.7) |>
  cols_label(first_quarter = "First detected") |>
  tab_options(table.font.size = 12)
```

## Top signals — neurological

```{r neuro}
neuro_top <- both |>
  filter(category == "Neurological") |>
  arrange(desc(eb05)) |>
  group_by(family) |>
  slice_head(n = 12) |>
  ungroup()

neuro_top |>
  select(family, vaccine, event, first_quarter, eb05, prr, ic, n_methods_flagged) |>
  gt() |>
  tab_header(title = "Top neurological signals",
             subtitle = "Statistics at first-detection quarter · top 12 per family by EB05") |>
  fmt_number(columns = c(eb05, prr, ic), decimals = 1) |>
  data_color(columns = eb05,
             palette = c("white", "#08519c"),
             alpha = 0.7) |>
  cols_label(first_quarter = "First detected") |>
  tab_options(table.font.size = 12)
```

## Top signals — thrombotic

```{r thrombotic}
thromb_top <- both |>
  filter(category == "Thrombotic") |>
  arrange(desc(eb05)) |>
  group_by(family) |>
  slice_head(n = 12) |>
  ungroup()

thromb_top |>
  select(family, vaccine, event, first_quarter, eb05, prr, ic, n_methods_flagged) |>
  gt() |>
  tab_header(title = "Top thrombotic signals",
             subtitle = "Statistics at first-detection quarter · top 12 per family by EB05") |>
  fmt_number(columns = c(eb05, prr, ic), decimals = 1) |>
  data_color(columns = eb05,
             palette = c("white", "#54278f"),
             alpha = 0.7) |>
  cols_label(first_quarter = "First detected") |>
  tab_options(table.font.size = 12)
```

## Bubble chart: EB05 vs methods-flagged, faceted by category

```{r bubble, fig.height=9}
plot_df <- both |>
  mutate(category = factor(category, levels = c("Cardiac", "Neurological", "Thrombotic")))

VACCINE_COLOURS <- c(
  "Pfizer-BioNTech"        = "#1a6b9a",
  "Moderna"                = "#e05f2a",
  "Janssen (J&J)"          = "#2e8b57",
  "Pfizer Bivalent"        = "#6baed6",
  "Moderna Bivalent"       = "#fd8d3c",
  "Novavax"                = "#756bb1",
  "Other / unknown"        = "#bdbdbd",
  "Shingrix (recombinant)" = "#a50f15",
  "Zostavax (live)"        = "#3690c0",
  "Zoster (no brand)"      = "#999999"
)

ggplot(plot_df,
       aes(x = n_methods_flagged, y = eb05,
           size = pmin(observed, 5000), colour = vaccine)) +
  geom_jitter(width = 0.15, height = 0, alpha = 0.6) +
  scale_y_log10(labels = comma) +
  scale_size_continuous(range = c(1.5, 9), guide = "none") +
  scale_colour_manual(values = VACCINE_COLOURS) +
  facet_grid(category ~ family, scales = "free_y") +
  labs(
    title    = "Signal landscape — cardiac, neurological, thrombotic",
    subtitle = "Each point is a (vaccine, event) pair surviving the consensus filter",
    x        = "Number of methods flagged (of 4)",
    y        = "EB05 (log10 scale)",
    colour   = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom",
        strip.background = element_rect(fill = "#f0f0f0", colour = NA),
        strip.text       = element_text(face = "bold"))
```

## Signal emergence timeline

When did each signal first cross the consensus threshold? The chart below
shows the **first-detection quarter** for the top-strength signals in
each domain. Cluster patterns by quarter often track external events
(product launches, media attention, EUA / approval expansions).

```{r emergence, fig.height=10}
emergence_df <- both |>
  group_by(family) |>
  slice_max(eb05, n = 30, with_ties = FALSE) |>
  ungroup() |>
  mutate(
    event_short = str_trunc(event, 48),
    first_date  = as.Date(paste0(
      str_extract(first_quarter, "^\\d{4}"), "-",
      sprintf("%02d", (as.integer(str_extract(first_quarter, "\\d$")) - 1L) * 3L + 1L),
      "-01"
    ))
  ) |>
  filter(!is.na(first_date))

ggplot(emergence_df,
       aes(x = first_date,
           y = fct_reorder(event_short, first_date, .desc = TRUE),
           colour = vaccine,
           size   = eb05)) +
  geom_point(alpha = 0.8) +
  scale_colour_manual(values = VACCINE_COLOURS) +
  scale_size_continuous(range = c(2, 7), guide = "none") +
  scale_x_date(date_breaks = "6 months", date_labels = "%Y Q%q") +
  facet_grid(family ~ ., scales = "free_y", space = "free_y") +
  labs(title    = "First quarter each signal crossed the consensus threshold",
       subtitle = "Point size = EB05 at first detection · top 30 per family by EB05",
       x = NULL, y = NULL, colour = NULL) +
  theme_minimal(base_size = 11) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1),
        axis.text.y = element_text(size = 9),
        legend.position  = "bottom",
        strip.background = element_rect(fill = "#f0f0f0", colour = NA),
        strip.text       = element_text(face = "bold"))
```

## Manufacturer comparison

```{r manufacturer}
mfr <- both |>
  count(family, vaccine, category, name = "signals") |>
  group_by(family, vaccine) |>
  mutate(total = sum(signals)) |>
  ungroup()

ggplot(mfr, aes(x = reorder(vaccine, -total), y = signals, fill = category)) +
  geom_col() +
  scale_fill_manual(values = c("Cardiac" = "#a50f15",
                               "Neurological" = "#08519c",
                               "Thrombotic"   = "#54278f")) +
  facet_wrap(~ family, scales = "free_x") +
  labs(title    = "Per-vaccine signal counts in the three pre-specified domains",
       subtitle = "Stacked bars: cardiac + neurological + thrombotic · first-detection count",
       x = NULL, y = "Distinct signals (first-detection quarter)") +
  theme_minimal(base_size = 12) +
  theme(axis.text.x  = element_text(angle = 35, hjust = 1),
        legend.title = element_blank())
```

## Interpretation

A few patterns worth flagging:

1. **COVID-19 cardiac signals are dominated by myocarditis/pericarditis
   in younger reporters and by mRNA products** — consistent with the
   epidemiological signal that emerged in 2021 and that subsequently
   appeared in product labelling. The mRNA bivalent boosters carry a
   smaller but visible cardiac signal in the same direction.

2. **The Janssen viral-vector product's thrombotic signature — TTS /
   cerebral venous sinus thrombosis** — survives the consensus filter
   with very high EB05 and remains the dominant product-specific
   thrombotic signal in the dataset. This is the signal that led to the
   restriction and eventual withdrawal of Janssen.

3. **Shingles vaccine signals in the three target domains are
   substantially sparser than COVID's.** That is expected: shingles
   vaccines (especially recombinant Shingrix) have a different reaction
   profile dominated by local and inflammatory reactions, and the
   disease itself causes post-herpetic neuralgia which can confound any
   "neurological" interpretation (the vaccine is given to prevent the
   condition that produces those events). Cardiac and thrombotic
   signals for shingles vaccines, where they appear, are typically at
   lower EB05 with smaller observed counts.

4. **Zostavax (live) shows a higher per-product signal density than
   Shingrix (recombinant) in these domains.** This is consistent with
   the live-attenuated platform's broader systemic reactogenicity and
   the older age distribution of recipients in the era when Zostavax
   was the only option.

5. **Differences between products within a family matter more than the
   family-level summary.** Pooling all COVID-19 vaccines into a single
   row obscures the mRNA-vs-viral-vector pattern that drove regulatory
   action. Pooling the two shingles products would obscure the
   live-vs-recombinant difference. The bubble chart above is the
   appropriate granularity.

::: callout-warning
**Confounding by indication is a real risk for shingles signals.**
"Post-herpetic neuralgia" is both a neurological event in VAERS and the
exact outcome shingles vaccination is meant to prevent. Any signal
involving herpes-zoster sequelae must be read with that bias in mind.
:::

## Limitations

- VAERS is a passive, voluntary reporting system. Reporting rates vary
  by media attention, secular trends, and demographic shifts.
- Disproportionality detects deviation from the *reporting* baseline,
  not from the *true clinical* baseline. Differential reporting can
  produce signals; differential clinical occurrence may not.
- The three event categories above are pre-specified for this
  reanalysis; broader categories (e.g., allergic, gastrointestinal,
  hepatic, renal) are excluded by design.
- No demographic stratification (age, sex, dose number) is performed
  here. Stratification typically increases sensitivity but is out of
  scope for a top-line reanalysis.
- All four methods rely on the same 2×2 contingency. They are not
  statistically independent; agreement across methods means agreement
  about the count distribution, not independent confirmation.

## Replicable R

The full source of this article is the Quarto file
`shingles_vaccine_analysis.qmd` in the
[globalpatientsafety](https://github.com/harlananelson) repository. The
`safetysignal` R package implementing the four detection methods is
similarly open-source. All input data is CDC VAERS, available without
restriction at <https://vaers.hhs.gov/data.html>.

```{r session}
sessionInfo()
```
