---
title: "The Liver Is the Limit: Elevidys's Boxed Warning Is an AAV Class Effect"
subtitle: "Reading the fatal-hepatotoxicity warning against the whole gene-therapy class — and finding the signal tracks the delivery route, not the product"
author: "Global Patient Safety"
date: 2026-07-21
categories: [FAERS, pharmacovigilance, gene therapy, hepatotoxicity, class effect]
format:
  html:
    toc: true
    toc-depth: 3
    theme: flatly
    embed-resources: true
execute:
  echo: false
  warning: false
  message: false
params:
  data_path: "/srv/shiny-server/gps-patient/data/signals_faers.parquet"
---

```{r setup}
#| include: false
library(arrow)
library(dplyr)
library(gt)

data_candidates <- c(
  params$data_path,
  "/home/harlan/projects/gps-patient/data/signals_faers.parquet",
  "/home/harlan/data/signal-compute/signals_faers_v2026-07-08.parquet"
)
data_path <- data_candidates[file.exists(data_candidates)][1]
if (is.na(data_path)) stop("signals_faers.parquet not found")
d <- open_dataset(data_path) |> collect()
LQ <- max(d$quarter)

# Approved AAV / gene-therapy products, each with its delivery route — the axis
# the analysis turns on. "systemic" = IV infusion (the vector reaches the liver);
# "local" / "ex-vivo" = the liver is never systemically exposed.
prods <- tibble::tribble(
  ~name,        ~regex,                           ~indication,        ~route,
  "Elevidys",   "elevidys|delandistrogene",       "DMD",              "IV — systemic",
  "Zolgensma",  "zolgensma|onasemnogene",          "SMA (infants)",   "IV — systemic",
  "Hemgenix",   "hemgenix|etranacogene",           "Haemophilia B",   "IV — liver-directed",
  "Roctavian",  "roctavian|valoctocogene",         "Haemophilia A",   "IV — liver-directed",
  "Luxturna",   "luxturna|voretigene",             "Retinal dystrophy","Subretinal — local",
  "Casgevy",    "casgevy|exagamglogene",           "SCD / β-thal",    "Ex-vivo edited cells"
)
HEP_RX <- paste0("hepatic failure|liver injury|hepatocellular|drug-induced liver|hepatotox",
                 "|transaminases|aminotransferase|hyperbilirubin|liver disorder|hepatitis acute",
                 "|jaundice|hepatic enzyme")

prod_stats <- function(rx) {
  tot <- d |> filter(grepl(rx, drug))
  hep <- tot |> filter(grepl(HEP_RX, event, ignore.case = TRUE))
  # Headline pair = strongest hepatic signal with a non-trivial count (observed >= 5),
  # so a single n=1 extreme EB05 can't stand in for a product's hepatic profile.
  best <- hep |> filter(observed >= 5) |> arrange(desc(eb05)) |> slice(1)
  tibble::tibble(
    total_reports = sum(tot$observed),
    hep_reports   = sum(hep$observed),
    max_eb05      = if (nrow(best)) round(best$eb05, 0) else NA_real_,
    max_flags     = if (nrow(best)) best$n_methods_flagged else 0L
  )
}
class_tbl <- prods |> rowwise() |> mutate(s = list(prod_stats(regex))) |>
  mutate(total_reports = s$total_reports, hep_reports = s$hep_reports,
         max_eb05 = s$max_eb05, max_flags = s$max_flags) |> ungroup() |> select(-s, -regex)

# Elevidys inline values.
el <- function(pt, q = LQ) d |> filter(grepl("elevidys|delandistrogene", drug),
  tolower(event) == tolower(pt), quarter == q) |> arrange(desc(eb05)) |> slice(1)
alt <- el("Alanine aminotransferase increased")
alf <- el("Acute hepatic failure")
```

On **November 14, 2025**, the FDA added a **boxed warning** for acute serious liver injury and
acute liver failure to **Elevidys** (delandistrogene moxeparvovec), the AAV gene therapy for
Duchenne muscular dystrophy, and removed its non-ambulatory indication — after **two pediatric
deaths** from acute liver failure. Hepatotoxicity is a recognized hazard of AAV gene therapy, so
the natural regulatory framing is product-specific: a warning on *this* drug.

This post asks a different question, and it is one no single-product warning can answer: **is
the liver signal specific to Elevidys, or is it a property of the whole class?** The answer in
FAERS is unusually clean — and it points not at a molecule but at a **delivery route**.

::: {.callout-important}
## What this can and cannot show

Disproportionality compares how often a drug–event pair is reported to what the rest of the
database predicts. It cannot establish causation or count harmed patients, and gene-therapy
products are low-volume, so several estimates below rest on small numbers (the EB05 lower bound
accounts for that, and every count is shown). These products also treat different diseases in
different populations with different co-medication — confounding that the discussion addresses
head-on rather than assuming away.
:::

## The mechanism, and why route is the right axis

Most systemically infused (intravenous) AAV vectors are cleared predominantly by the **liver** —
hepatocytes are the principal sink for circulating AAV, regardless of the therapeutic target.
The dose-limiting toxicity of systemic AAV is therefore hepatocellular: transaminase
elevations, immune-mediated hepatocellular injury, and, at the extreme, acute liver failure.
This predicts a specific pattern. Gene therapies delivered **into the bloodstream** should share
a hepatic signal — even when their indications and targets differ completely — while therapies
that **never expose the liver systemically** (a local injection, or cells edited outside the
body and re-infused) should not.

The approved products split cleanly along that axis, which is why the table below is organised
by route, not by manufacturer.

## The class league table

```{r class-table}
class_tbl |>
  mutate(`Hepatic signal` = ifelse(is.na(max_eb05), "none",
           paste0("EB05 ", formatC(max_eb05, format = "d"), " (", max_flags, "/4)")),
         `Total reports` = formatC(total_reports, format = "d", big.mark = ","),
         `Hepatic reports` = formatC(hep_reports, format = "d", big.mark = ",")) |>
  select(Product = name, Indication = indication, Route = route,
         `Total reports`, `Hepatic reports`, `Hepatic signal`) |>
  gt() |>
  tab_header(title = md("**Hepatic disproportionality across the approved gene-therapy class**"),
             subtitle = md("Hepatic PT cluster: liver injury / failure, transaminase & bilirubin elevations, DILI, hepatotoxicity. Strongest single hepatic pair per product.")) |>
  tab_source_note(md("*Hepatic signal = the highest EB05 hepatic pair for that product and the number of the four methods flagging it. 'none' = zero hepatic reports in the dataset.*"))
```

The split is complete:

- **Every intravenously infused product carries a hepatic signal.** Elevidys, Zolgensma,
  Hemgenix, and Roctavian — four products, four different diseases (DMD, spinal muscular
  atrophy, haemophilia B, haemophilia A), four different target tissues — all flag for
  hepatocellular injury, most with all four methods. Zolgensma's hepatic footprint is the
  largest by far (thousands of reports); it has carried a hepatotoxicity boxed warning and
  mandatory prophylactic corticosteroids since approval. The Elevidys warning is new; the class
  hazard is not.
- **The two products that never systemically expose the liver carry no hepatic signal at all.**
  **Luxturna** is injected *locally beneath the retina*; **Casgevy** is not an in-vivo AAV at all
  but autologous stem cells edited *outside the body* and re-infused. Both have substantial
  report counts — and **zero** hepatic reports. They are the built-in negative controls, and
  they behave exactly as the route hypothesis predicts.

That a hepatic signal appears across four unrelated indications but vanishes in precisely the two
products that bypass systemic liver exposure is hard to explain as confounding by indication. It
is much more simply explained by where the vector goes.

## Elevidys up close: the injury spectrum is loud, the fatal endpoint is faint

Within Elevidys, the hepatic signal is dominated by the **injury spectrum** — the enzyme
elevations and hepatocellular-injury terms that precede failure. In `r LQ`, alanine
aminotransferase increased shows **`r alt$observed` reports against `r formatC(alt$expected, format="f", digits=2)` expected (EB05 `r formatC(alt$eb05, format="f", digits=1)`, all `r alt$n_methods_flagged` methods)**, and this spectrum — ALT/AST increased, transaminases increased, liver injury, drug-induced liver injury, hepatotoxicity — flags with all four methods across every recent quarter.

The endpoint the boxed warning is *named* for behaves differently. **Acute hepatic failure**
appears at just **`r alf$observed` reports** in `r LQ` (EB05 `r formatC(alf$eb05, format="f", digits=1)`) — barely at threshold. This is the same lesson the carbidopa/levodopa
analysis reached from the other direction: **the rare, severe outcome is nearly invisible in
disproportionality, while the mechanistic precursor is unmistakable.** A warning built on two
deaths will not light up a database the way the thousands of upstream enzyme elevations do — and
the enzyme elevations are exactly what a monitoring programme should watch.

## Reading it honestly

The disciplined conclusion:

- **Elevidys's hepatotoxicity is real and consistent** in FAERS — but it is **not distinctive**.
  It is one instance of a hazard shared by every systemically infused AAV gene therapy.
- **The signal tracks the delivery route**, not the product: it is present wherever the vector
  reaches the liver and absent where it does not. The regulatory action on one product is best
  read as a class-wide, biodistribution-driven risk that monitoring and prophylaxis (as already
  standard for Zolgensma) are designed to manage.
- **The fatal endpoint is not the detectable signal.** Transaminase elevation is — which is
  reassuring, because it is measurable and actionable before failure.

::: {.callout-note}
## Limitations

- **Small numbers.** Hemgenix, Roctavian, and Elevidys's fatal-endpoint counts are small; the
  EB05 lower bound already shrinks toward the null to reflect this, and all counts are shown.
- **Confounding by indication and co-medication.** Populations and regimens differ enormously
  (infants on prophylactic steroids for Zolgensma; ambulatory boys for Elevidys). But the hepatic
  signal's *consistency across* these very different settings, and its *absence* in the two
  liver-sparing products, is what makes the route interpretation more plausible than a per-drug
  artifact — not less.
- **Known-effect / notoriety reporting.** Zolgensma's warning is years old and Elevidys's is
  recent; some reporting is stimulated by the warnings themselves.
- **Passive reports are not incidence**, and disproportionality is not causation.
:::

## Methods & data

FAERS through `r LQ`, four-quarter rolling windows, four methods (GPS/EBGM EB05, PRR, ROR,
BCPNN/IC), signal rule EB05 ≥ 2.0 with ≥ 2 of 4. Products identified by brand + INN regex;
the hepatic Preferred-Term cluster is listed above. The strongest hepatic pair per product is
tabulated; Elevidys is examined term by term. Data: FDA Adverse Event Monitoring System (AEMS)
quarterly extract, continuing the FAERS series.

## Sources

- FDA, "FDA Takes Action on New Boxed Warning for Acute Serious Liver Injury and Acute Liver Failure Following Treatment with Elevidys and Revised Indication" (Nov 14, 2025). <https://www.fda.gov/vaccines-blood-biologics/safety-availability-biologics/fda-takes-action-new-boxed-warning-acute-serious-liver-injury-and-acute-liver-failure-following>
- AAP News, "After 2 child deaths, FDA approves updated safety label for Duchenne muscular dystrophy treatment." <https://publications.aap.org/aapnews/news/33714/>
- LiverTox (NCBI Bookshelf), "Delandistrogene Moxeparvovec." <https://www.ncbi.nlm.nih.gov/books/NBK618387/>
- Zolgensma US Prescribing Information (boxed warning: acute serious liver injury; prophylactic corticosteroids). <https://www.fda.gov/vaccines-blood-biologics/zolgensma>
- Data: FDA AEMS quarterly extract files (continuing FAERS). <https://fis.fda.gov/extensions/FPD-QDE-FAERS/FPD-QDE-FAERS.html>
