---
title: "Inside the AEMS Data: What 4.8 Million Reports Do and Don't Tell You"
subtitle: "The FDA's new Adverse Event Monitoring System, the quarterly data behind our signal analyses, and why the strongest 'signal' is usually an artifact"
author: "Global Patient Safety"
date: 2026-07-14
format:
  html:
    toc: true
    toc-depth: 3
    theme: flatly
    embed-resources: true
execute:
  echo: false
  warning: false
  message: false
params:
  data_path: "/srv/shiny-server/gps-patient/data/signals_faers.parquet"
---

```{r setup}
#| include: false
library(arrow)
library(dplyr)
library(gt)

data_candidates <- c(
  params$data_path,
  "/home/harlan/projects/gps-patient/data/signals_faers.parquet",
  "/home/harlan/data/signal-compute/signals_faers_v2026-07-08.parquet"
)
data_path <- data_candidates[file.exists(data_candidates)][1]
if (is.na(data_path)) {
  stop("signals_faers.parquet not found at any candidate path:\n  ",
       paste(data_candidates, collapse = "\n  "))
}
d <- open_dataset(data_path) |> collect()

LQ <- max(d$quarter)
q  <- d |> filter(quarter == LQ)

fmt <- function(x) formatC(x, format = "d", big.mark = ",")
n_pairs     <- nrow(q)
tot_reports <- sum(q$observed)
n_consensus <- sum(q$n_methods_flagged >= 2)
n_all4      <- sum(q$n_methods_flagged == 4)

# Live lookup for the annotated "cautionary" table: named drug+event pairs that
# top the raw EB05 leaderboard but are artifacts, each with a plain-language
# reason. Values are pulled live so they can't drift from the data.
caution <- tibble::tribble(
  ~drug,          ~event,                            ~artifact,          ~why,
  "endari",       "Sickle cell anaemia with crisis", "Indication",       "Endari (L-glutamine) is prescribed *for* sickle cell disease — so its reports are full of sickle cell crises. The drug treats the condition it appears to 'signal' for.",
  "gold bond original strength", "Mesothelioma",     "Litigation cluster","A wave of talc-related legal reporting, not a pharmacovigilance discovery. Mass-tort campaigns generate synchronized report bursts unrelated to new biology.",
  "nirsevimab",   "Respiratory syncytial virus bronchiolitis", "Indication / route", "Nirsevimab is given to *prevent* RSV disease in infants; RSV terms saturate its reports by construction.",
  "heparin",      "Heparin-induced thrombocytopenia","Definitional",     "A real, well-known reaction — but an *eponymous* one already in every label for decades. Disproportionality 'rediscovering' it is confirmation, not signal."
)
clook <- function(dr, ev) d |> filter(drug == dr, event == ev, quarter == LQ) |>
  arrange(desc(eb05)) |> slice(1)
```

## What AEMS is

On **March 11, 2026** the FDA launched the **Adverse Event Monitoring System (AEMS)**, a single
real-time public platform that consolidates the agency's previously separate adverse-event
databases — **FAERS** (drugs and biologics), **VAERS** (vaccines), and the animal AERS — with
devices, food, supplements, and tobacco folding in through 2026. The stated reasons were cost
(the legacy systems ran ~\$37M/year) and fragmentation; the new system publishes in real time
rather than quarterly and adds an AI-assisted dashboard.

For an independent analyst, the important fact is quieter and reassuring: **the bulk quarterly
data-extract files continue.** They are still published, still free, still in the same ASCII/XML
format that begins in 2004 — now under the AEMS name, with the January–March 2026 quarter posted
on April 28, 2026. Those extracts are the raw material behind every analysis on this site. This
page is a look *at the data itself*: how much of it there is, and how to read it without fooling
yourself.

## The current data at a glance

The most recent quarter in our processed dataset is **`r LQ`**. Within its four-quarter rolling
window:

```{r snapshot}
tibble::tibble(
  Metric = c("Distinct drug–event pairs",
             "Total reports (sum of observed counts)",
             "Pairs flagged by ≥ 2 of 4 methods (our signal rule)",
             "Pairs flagged by all 4 methods"),
  Value = c(fmt(n_pairs), fmt(tot_reports), fmt(n_consensus), fmt(n_all4))
) |>
  gt() |>
  tab_header(title = md(paste0("**AEMS / FAERS signals dataset — ", LQ, "**"))) |>
  cols_align("left", columns = Metric) |>
  cols_align("right", columns = Value)
```

Every one of those `r fmt(n_pairs)` pairs is scored by four disproportionality methods — the
Gamma-Poisson Shrinker (GPS/EBGM), the Proportional Reporting Ratio (PRR), the Reporting Odds
Ratio (ROR), and the BCPNN Information Component (IC). A pair becomes a "signal," in our fixed
convention, when its EB05 is at least 2.0 **and** at least two of the four methods agree.

Here is how the whole quarter distributes across method agreement:

```{r agreement}
q |>
  count(`Methods flagged` = n_methods_flagged) |>
  arrange(`Methods flagged`) |>
  mutate(Pairs = fmt(n),
         Share = paste0(formatC(100 * n / sum(n), format = "f", digits = 1), "%")) |>
  select(`Methods flagged`, Pairs, Share) |>
  gt() |>
  tab_header(title = md("**How many methods agree, across every pair this quarter**")) |>
  tab_source_note(md("*Most pairs are flagged by no method at all. Agreement by all four is the minority — and, as the next section shows, even that is not enough on its own.*"))
```

## Why we don't publish a "top signals" leaderboard

It is tempting to sort those `r fmt(n_all4)` four-method pairs by EB05 and call the top of the
list "the biggest safety signals." **Do not trust that list.** The strongest disproportionality
scores are dominated not by discoveries but by *artifacts* — and looking at the actual top of
the current quarter makes the point better than any warning:

```{r caution}
caution |>
  rowwise() |>
  mutate(r = list(clook(drug, event))) |>
  mutate(Observed = if (nrow(r)) r$observed else NA_integer_,
         EB05     = if (nrow(r)) round(r$eb05, 0) else NA_real_) |>
  ungroup() |>
  transmute(Drug = drug, Event = event, Observed, EB05,
            `Why it's not a discovery` = paste0("**", artifact, ".** ", why)) |>
  gt() |>
  fmt_markdown(columns = `Why it's not a discovery`) |>
  tab_header(title = md("**Four of the strongest 'signals' this quarter — and why each is an artifact**")) |>
  tab_source_note(md("*EB05 and observed counts are pulled live from the current data; the annotations are fixed.*"))
```

Each of these clears every statistical bar — high EB05, all four methods, hundreds of reports —
and each is meaningless as a *discovery*. They illustrate the four ways a raw leaderboard lies:

- **Confounding by indication** — a drug's reports are full of the disease it treats.
- **Litigation and media clusters** — synchronized reporting campaigns, not biology.
- **Route and administration effects** — the reports describe how the drug is given.
- **Definitional / known reactions** — real, but already on the label for years.

A disproportionality score is a *question generator*, never an answer. The number tells you where
to look; it cannot tell you what you will find.

## Analysis done right

Useful signal detection starts from a **specific, pre-registered question** and spends most of
its effort on confounding, not on the score. Two pieces on this site show that pattern on
published work (not a raw leaderboard):

- [**Signal & Noise**](/methods) — the house methods field guide: one real signal, one trap, one
  fix at a time; fixed rule of EB05 ≥ 2 with ≥ 2 of 4 methods agreeing.
- [**Tested Against VAERS**](/christine_cotton) — Christine Cotton's trial-era safety categories
  checked against the post-market VAERS record with the same four-method consensus filter.

Analyses on this site present event names as they appear in the public FAERS/VAERS releases
(Preferred Term labels in those extracts). We do not redistribute MedDRA terminology files or
hierarchy products.

## Explore it yourself

To search the full dataset interactively — any drug, any event, signal trajectories over
quarters — use **[faers.mobi](https://faers.mobi)**, which runs the same four-method engine on
the same AEMS/FAERS extract with server-side search across every drug–event pair.

::: {.callout-note}
## How to read anything on this page

These are counts of *reports*, not counts of *patients harmed*. Adverse-event reporting is
voluntary, unverified, and shaped by attention and litigation. Disproportionality compares a
pair's reporting rate to the rest of the database; it **cannot** establish causation, and a
high score is a reason to investigate, not a conclusion. Aggregate counts like those above carry
no patient-identifying information.
:::

## Methods & data

Data: FDA Adverse Event Monitoring System (AEMS) quarterly extract files, continuing the FAERS
series (<https://fis.fda.gov/extensions/FPD-QDE-FAERS/FPD-QDE-FAERS.html>), processed into
four-quarter rolling windows and scored by GPS/EBGM (EB05 posterior lower bound), PRR, ROR, and
BCPNN/IC. Signal rule: EB05 ≥ 2.0 with ≥ 2 of 4 methods, applied uniformly. Figures on this page
are computed live from the current processed dataset (latest quarter: `r LQ`).

Sources: FDA, "FDA Launches New Adverse Event Look-Up Tool"
(<https://www.fda.gov/news-events/press-announcements/fda-launches-new-adverse-event-look-tool>);
CIDRAP, "FDA announces AEMS, new adverse-event database to replace VAERS"
(<https://www.cidrap.umn.edu/public-health/fda-announces-aems-new-adverse-event-database-replace-vaers>).
