---
title: "A Warning Built on 14 Cases: Does FAERS See the Carbidopa/Levodopa B6-Seizure Signal?"
subtitle: "The seizure the FDA warned about is invisible in disproportionality — but the mechanism behind it is one of the drug's loudest, most stable signals"
author: "Global Patient Safety"
date: 2026-07-12
categories: [FAERS, pharmacovigilance, neurology, drug safety, mechanism]
format:
  html:
    toc: true
    toc-depth: 3
    theme: flatly
    embed-resources: true   # self-contained HTML (matches christine-cotton/shingles); build_static_site.R ships only the single <id>.html
execute:
  echo: false
  warning: false
  message: false
params:
  # Default = server path (where this renders for publication). The setup chunk
  # falls back to the local workstation copy automatically if the server path is
  # absent, so the same file previews locally and deploys without editing.
  # Override with: quarto render … -P data_path:/some/other/signals_faers.parquet
  data_path: "/srv/shiny-server/gps-patient/data/signals_faers.parquet"
---

```{r setup}
#| include: false
library(arrow)
library(dplyr)
library(gt)
# Trajectory uses base R graphics (no ggplot2 dependency) so the article renders
# in the same environment as the other site articles (christine-cotton uses gt only).

# Resolve the FAERS signals parquet: prefer params$data_path (server default),
# fall back to the local home-dir copy when rendering on the workstation.
data_candidates <- c(
  params$data_path,
  "/home/harlan/projects/gps-patient/data/signals_faers.parquet",
  "/home/harlan/data/signal-compute/signals_faers_v2026-07-08.parquet"
)
data_path <- data_candidates[file.exists(data_candidates)][1]
if (is.na(data_path)) {
  stop("signals_faers.parquet not found at any candidate path:\n  ",
       paste(data_candidates, collapse = "\n  "))
}

d <- open_dataset(data_path) |> collect()

# Carbidopa/levodopa product family (the FDA-warned products), including the
# extended-release, enteral-suspension, and foscarbidopa/foslevodopa prodrug
# formulations. FAERS free-text drug names fragment badly, so we match a regex
# family and show the per-formulation breakdown rather than a single string.
DRUG_RX <- "carbidopa|sinemet|rytary|duopa|stalevo|foscarbidopa|levodopa"
cl <- d |> filter(grepl(DRUG_RX, drug, ignore.case = TRUE))

# The largest, most interpretable string — the plain generic. Used as the
# reference formulation for prose figures; other formulations corroborate.
REF <- "carbidopa levodopa"

# Analysis convention: report the MOST RECENT quarter as the current snapshot,
# not the maximum EB05 across all quarters. For a rare-event drug, taking the
# single most extreme quarter across a 32-quarter panel cherry-picks noise; the
# latest rolling window is the honest "where do we stand now" statistic. Observed
# counts are shown alongside every EB05 so the reader can weigh the evidence.
LATEST_Q <- max(cl$quarter)
EB_THRESHOLD <- 2.0   # EB05 >= 2.0 with >=2 of 4 methods flagging — fixed, no per-event tuning.

# Reference-formulation signal for one PT in the latest quarter.
ref_row <- function(pt_exact) {
  cl |>
    filter(drug == REF, quarter == LATEST_Q,
           tolower(event) == tolower(pt_exact)) |>
    slice(1)
}
ebv    <- function(pt) formatC(ref_row(pt)$eb05,   format = "f", digits = 1)
icv    <- function(pt) formatC(ref_row(pt)$ic025,  format = "f", digits = 2)
obsv   <- function(pt) ref_row(pt)$observed
expv   <- function(pt) formatC(ref_row(pt)$expected, format = "f", digits = 2)
flagv  <- function(pt) ref_row(pt)$n_methods_flagged

# Pooled observed across all formulations for a PT in the latest quarter.
pooled_obs <- function(pt_exact) {
  cl |> filter(quarter == LATEST_Q, tolower(event) == tolower(pt_exact)) |>
    summarise(n = sum(observed)) |> pull(n)
}
```

On **March 20, 2026**, the FDA announced it is *requiring* a new warning on every
carbidopa/levodopa product: the drug can cause **vitamin B6 (pyridoxine) deficiency and
B6-deficiency–associated seizures**. The action rested on **14 seizure cases** identified in
the agency's own post-market reporting data, with the agency noting that the true number is
likely higher because of under-reporting.

Fourteen cases is a small number to hang a required label change on. It is also exactly the
situation where an independent pass over the *entire* reporting database is informative — it
can either corroborate the regulator's small series or show that the signal is invisible at
population scale. This post does that pass, using FAERS through `r LATEST_Q` and the site's
standard four-method disproportionality engine (Gamma-Poisson Shrinker / EBGM, PRR, ROR, and
BCPNN's Information Component), with a single fixed rule: **EB05 ≥ 2.0 and at least two of the
four methods flagging**, decided before looking at which events appeared.

The result is more interesting than a simple yes/no.

::: {.callout-important}
## What this can and cannot show

Disproportionality asks one narrow question: *is a drug–event pair reported more often than
the rest of the database would predict?* It **cannot** establish causation, it cannot count
how many people were harmed, and it is shaped by who reports and why. A signal is a reason to
look harder, not a verdict. Confounding by indication is real and central here: Parkinson's
disease itself carries elevated seizure and neuropathy risk, so a raw signal for a Parkinson's
drug cannot be cleanly separated from the disease it treats. Nothing below is a claim that
carbidopa/levodopa *causes* any specific patient's event. The point is narrower — whether the
*categories* the FDA warning names show disproportionate reporting, and whether they do so in
a way that is *mechanistically coherent*.
:::

## The mechanism the warning invokes

The warning is unusual in naming a concrete biochemical mechanism, which makes it testable.
Levodopa is metabolized in the periphery in reactions that consume pyridoxal-5′-phosphate, the
active form of vitamin B6. Carbidopa — the decarboxylase inhibitor added to keep levodopa from
being broken down before it reaches the brain — is a **hydrazine derivative** that reacts
directly with pyridoxal-5′-phosphate and inactivates it. The two effects compound into a
functional B6 deficiency. Because B6 is a cofactor in GABA synthesis, severe deficiency lowers
the seizure threshold — the pathway the FDA warning describes: **drug → B6 depletion →
(peripheral neuropathy, and, at the extreme) seizures**.

That gives us a chain of Preferred Terms to test, not just the headline outcome:
*Vitamin B6 deficiency* and *Blood homocysteine increased* (the metabolic derangement),
*Peripheral neuropathy* and *Polyneuropathy* (the common downstream injury), and *Seizure* /
*Generalised tonic-clonic seizure* (the rare, severe endpoint the warning is named for).

## What FAERS shows

### The seizure signal is essentially absent at population scale

Here is the current-quarter (`r LATEST_Q`) picture for **Seizure**, broken out by
formulation so the free-text fragmentation is visible rather than hidden:

```{r seizure-table}
cl |>
  filter(quarter == LATEST_Q, event == "Seizure") |>
  arrange(desc(observed)) |>
  transmute(Formulation = drug, Observed = observed,
            Expected = round(expected, 1), EB05 = round(eb05, 2),
            PRR = round(prr, 1), ROR = round(ror, 1),
            `Methods flagged` = n_methods_flagged) |>
  gt() |>
  tab_header(title = md("**Seizure** — carbidopa/levodopa formulations, latest FAERS quarter"),
             subtitle = md(paste0("Quarter: ", LATEST_Q,
                                  ". Signal rule: EB05 ≥ 2.0 with ≥2 of 4 methods."))) |>
  tab_source_note(md("*EB05 is the conservative lower bound of the Bayesian (GPS/EBGM) posterior. Expected is the count predicted if the drug and event were independent.*"))
```

For the two highest-volume formulations, **observed reports of seizure are at or below what
independence predicts** — for `r REF`, `r obsv("Seizure")` observed against
`r expv("Seizure")` expected, an EB05 of `r ebv("Seizure")`, with `r flagv("Seizure")` of four
methods flagging. Only the smaller *Sinemet* and *Stalevo* strings clear the threshold, on a
handful of reports. Pooled across all formulations, plain *Seizure* has hovered near or below
the signal line for years. **In other words: the outcome the warning is named for does not
present as a disproportionality signal.** That is not evidence the warning is wrong — it is
what you expect when a regulator acts on 14 cases of a *rare* complication. Rare outcomes are
precisely the ones disproportionality is worst at seeing.

### The mechanism, by contrast, is loud and stable

Now the same drug, same database, same rule — for **Vitamin B6 deficiency**:

```{r b6-table}
cl |>
  filter(quarter == LATEST_Q, event == "Vitamin B6 deficiency") |>
  arrange(desc(observed)) |>
  transmute(Formulation = drug, Observed = observed,
            Expected = round(expected, 2), EB05 = round(eb05, 1),
            IC025 = round(ic025, 2),
            `Methods flagged` = n_methods_flagged) |>
  gt() |>
  tab_header(title = md("**Vitamin B6 deficiency** — carbidopa/levodopa formulations, latest FAERS quarter"),
             subtitle = md(paste0("Quarter: ", LATEST_Q, "."))) |>
  tab_source_note(md("*IC025 is the lower bound of the BCPNN Information Component; > 0 is the conventional signal criterion.*"))
```

For `r REF`, vitamin B6 deficiency shows **`r obsv("Vitamin B6 deficiency")` observed reports
against `r expv("Vitamin B6 deficiency")` expected — an EB05 of
`r ebv("Vitamin B6 deficiency")` and an IC025 of `r icv("Vitamin B6 deficiency")`, with all
`r flagv("Vitamin B6 deficiency")` methods flagging.** Pooled across formulations there are
`r pooled_obs("Vitamin B6 deficiency")` such reports in this single quarter. An EB05 above 70
means the *conservative lower bound* of the Bayesian estimate is more than seventy-fold; this
is not a marginal signal.

And it is not a one-quarter fluke. The trajectory below shows the reference formulation's B6
deficiency signal quarter by quarter:

```{r b6-trajectory}
#| fig-width: 8
#| fig-height: 4
#| fig-alt: "Line plot of EB05 for carbidopa/levodopa and Vitamin B6 deficiency by quarter, sustained well above the signal threshold from 2022 through the latest quarter."
traj <- cl |>
  filter(drug == REF, event == "Vitamin B6 deficiency") |>
  mutate(qdate = as.Date(paste0(substr(quarter, 1, 4), "-",
                                (as.integer(substr(quarter, 6, 6)) - 1) * 3 + 1, "-01"))) |>
  arrange(qdate)

xr <- range(traj$qdate)
par(mar = c(4, 4, 3, 1))
plot(traj$qdate, traj$eb05, type = "n", log = "y", xaxt = "n",
     xlab = "", ylab = "EB05 (log scale)",
     main = "Carbidopa/levodopa — Vitamin B6 deficiency signal by quarter")
axis.Date(1, at = seq(xr[1], xr[2], by = "1 year"), format = "%Y")
abline(h = EB_THRESHOLD, lty = 2, col = "grey50")
text(xr[1], EB_THRESHOLD, "signal threshold (EB05 = 2)", pos = 3, adj = 0, cex = 0.75, col = "grey40")
lines(traj$qdate, traj$eb05, col = "#2c7fb8", lwd = 2)
# Point size scaled to observed report count in each 4-quarter window.
points(traj$qdate, traj$eb05, col = "#2c7fb8", pch = 19,
       cex = 0.6 + 2.4 * (traj$observed - min(traj$observed)) /
             (max(traj$observed) - min(traj$observed) + 1e-9))
mtext(paste0("Reference formulation '", REF, "'. Point size = observed reports."),
      side = 3, line = 0.2, cex = 0.8)
```

The signal has been present and above threshold, with all four methods agreeing, across every
recent quarter — **years before the FDA's March 2026 warning.** The disproportionality the
label change implies was already sitting in the public data.

### Neuropathy: a weaker, formulation-dependent echo

The intermediate step in the mechanism — nerve injury from B6 depletion — shows up more
faintly and unevenly. *Polyneuropathy* clears the threshold on the continuous-infusion
formulations (a plausible dose/exposure story: steadier, higher levodopa turnover), while
plain *Peripheral neuropathy* for the oral generic is reported *below* expectation. The
neuropathy picture is real but modest, and it does not carry the article's weight — the two
findings that do are the **absence** of a seizure signal and the **stability** of the B6
deficiency signal.

## Reading it honestly

Put the two results side by side and the shape is coherent, not contradictory:

- The **rare, severe endpoint** (seizure) — the one the warning is named for — is invisible in
  disproportionality. That is the expected behavior of a method that needs volume, applied to
  an outcome the FDA itself found only 14 times.
- The **upstream mechanism** (vitamin B6 deficiency) — which no one would report unless a
  clinician specifically measured it — is one of the strongest and most durable signals in the
  drug's entire FAERS profile.

The honest reading is that FAERS *does* corroborate the biology of the warning, but through the
mechanism rather than the outcome. You cannot see the seizures in the aggregate data; you can
see, unmistakably and for years, the metabolic derangement that the warning says produces them.
For a signal-detection method, that is a useful lesson in where to look: when the named outcome
is rare, the mechanistic precursor can be the visible signal.

::: {.callout-note}
## Limitations

- **Confounding by indication.** Parkinson's disease independently raises seizure and neuropathy
  risk; disproportionality cannot separate drug from disease. An active-comparator analysis
  (against other Parkinson's therapies) would sharpen this and is the natural follow-up.
- **Small numbers.** Even the strong B6 signal rests on ~10–20 reports per quarter. The EB05
  lower bound already accounts for this (that is what shrinkage does), but the absolute counts
  are small and every figure here is shown with its observed count for that reason.
- **Drug-name fragmentation.** FAERS free-text splits one drug across dozens of strings; we
  report per-formulation and use the largest generic as reference rather than pooling the
  contingency tables. A properly pooled re-computation would likely *strengthen* the B6 signal,
  not weaken it.
- **Reporting dynamics.** Publicity around the warning will stimulate reporting. Our data ends
  at `r LATEST_Q`, *before* the March 2026 action, so the signal here is not an artifact of the
  warning itself. The post-warning quarters — now available under FDA's new AEMS extract, which
  continues the FAERS quarterly files — are the obvious next check for a stimulated-reporting
  spike.
- **Disproportionality is not causation**, and passive reports are not incidence.
:::

## Methods

FAERS reports through `r LATEST_Q`, four-quarter rolling windows. Each drug–event pair is
scored by four disproportionality methods computed independently: the Gamma-Poisson Shrinker
(GPS/EBGM, reported as the EB05 lower bound of the posterior credible interval), the
Proportional Reporting Ratio (PRR), the Reporting Odds Ratio (ROR), and the Bayesian Confidence
Propagation Neural Network Information Component (IC, reported as IC025). The pre-specified
signal rule is EB05 ≥ 2.0 with at least two of the four methods flagging, applied uniformly to
every Preferred Term with no per-event tuning. Carbidopa/levodopa products were identified by a
drug-name regex covering the oral generic, Sinemet, Rytary, Duopa, Stalevo, and the
foscarbidopa/foslevodopa prodrug; results are shown per formulation. Analysis is restricted to
the most recent quarter as the current snapshot, with the reference-formulation trajectory shown
across all quarters.

## Sources

- FDA Drug Safety Communication, "FDA Is Requiring a Warning about Vitamin B6 Deficiency and Associated Seizures for Drug Products Containing Carbidopa/Levodopa" (Mar 20, 2026). <https://www.fda.gov/drugs/drug-safety-communications/fda-requiring-warning-about-vitamin-b6-deficiency-and-associated-seizures-drug-products-containing>
- American Parkinson Disease Association, "FDA's New Warning: Link of Carbidopa/Levodopa to Vitamin B6 Deficiency and Seizures." <https://www.apdaparkinson.org/article/fdas-new-warning-link-of-carbidopa-levodopa-to-vitamin-b6-deficiency-and-seizures/>
- Evans SJW, Waller PC, Davis S. "Use of proportional reporting ratios (PRRs) for signal generation from spontaneous adverse drug reaction reports." *Pharmacoepidemiol Drug Saf.* 2001.
- DuMouchel W. "Bayesian data mining in large frequency tables, with an application to the FDA spontaneous reporting system." *The American Statistician.* 1999.
- Data: FDA Adverse Event Monitoring System (AEMS) quarterly extract files (continuing the FAERS series). <https://fis.fda.gov/extensions/FPD-QDE-FAERS/FPD-QDE-FAERS.html>
